---
title: "Understanding US Student Loan Debt"
description: |
  Perform data visualization and determine the most preferred method of payment for student loans in the US.
categories:
  - TidyTuesday 
  - Data viz

# draft: true
date: 2019-11-26
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```




```{r, message=FALSE, include=FALSE}
library(tidyverse)
library(extrafont)
library(directlabels)
library(ggrepel)
library(patchwork)
```


This week the data was inspired from the [Dignity & Debt](https://www.dignityanddebt.org/projects/student-debt-racial-disparities/) project. This will help in understanding and spreading awareness around Student Loan debt. The data source is [here](https://studentaid.gov/data-center/student/default)

The objective for this is to perform data visualization and determine the most preferred method of payment for student loans in the US. I will use `patchwork` package to combine plots.

Loading the data

```{r}
loans <- read.csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-11-26/loans.csv")

glimpse(loans)
```

Original data file can be accessed through the weekly TidyTuesday Github repository in this [link](https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-11-26/loans.csv) and data source is [here](https://studentaid.ed.gov/sa/about/data-center/student/default)



Renaming columns 

```{r}
loans %>% 
  mutate(year = case_when(year == 15 ~ 2015, 
                          year == 16 ~ 2016,
                          year == 17 ~ 2017,
                          year == 18 ~ 2018)) -> loans

```


```{r}
loans %>%
  group_by(year) %>%
  summarise(Total.starting = sum(starting, na.rm=TRUE)/10^9,
            Total.added = sum(added, na.rm = TRUE)/10^9,
            Total.repaid = sum(total, na.rm=TRUE)/10^9,
            consolidation.pay = sum(consolidation, na.rm = TRUE)/10^9,
            rehabilitation.pay = sum(rehabilitation, na.rm = TRUE)/10^9,
            voluntary.pay = sum(voluntary_payments, na.rm=TRUE)/10^9,
            wagegarnishment.pay =sum(wage_garnishments, na.rm = TRUE)/10^9 ) %>%
  arrange(desc(year))  %>%
  ungroup() -> payment.summary
```

```{r, layout="l-body-outset", preview=TRUE}
loan1 <- payment.summary %>% 
  pivot_longer(starts_with("Total"), names_to = "status", values_to = "totals")  %>% 
  mutate(label = if_else(year == max(year), as.character(status), NA_character_)) %>% 
  ggplot(mapping = aes(year, totals, col=status)) +
    geom_line(show.legend = FALSE)+
    scale_color_discrete(guide = "none") +
    theme_classic() +
  labs(title= "Loan Amounts per Year",
       x= "Year", y="Amount in USD (billion)", caption = "By am_innocenter for #TidyTuesdy 26.11.2019 \n Data source : studentaid.ed.gov")+
  geom_label_repel(aes(label=label),nudge_x = 1, na.rm = TRUE )+
  theme(panel.background = element_rect(fill="gray8"))  + #, colour = "#6D9EC1")) +
    theme(text = element_text( size = 12),
      plot.caption = element_text( size=10, color = "grey40"),
        plot.title =element_text(size = 14, face="bold") ) 
```




```{r}
loan2  <-  payment.summary %>% 
  pivot_longer(starts_with("Total"), names_to = "status", values_to = "totals")  %>% 
  mutate(label = if_else(year == max(year), as.character(status), NA_character_)) %>% 
  ggplot(mapping = aes(year, totals, col=status)) +
    geom_line(show.legend = FALSE)+
    scale_color_discrete(guide = "none") +
    scale_y_log10() +
    theme_classic() +
    labs(title= "Loan Amounts per Year", 
       caption = "By am_innocenter for #TidyTuesdy 26.11.2019 \n Data source : studentaid.ed.gov", 
       x= "Year", y="Log 10 of Amount in USD (billion)"
    )+
    geom_label_repel(aes(label=label),nudge_x = 1, na.rm = TRUE )+
    theme(panel.background = element_rect(fill="gray8"))+ #, colour = "#6D9EC1")) +
    theme(text = element_text( size = 12),
      plot.caption = element_text( size=10, color = "grey40"),
        plot.title =element_text(size = 14, face="bold") ) 
```


```{r}
loan1 + loan2
```




```{r}
payment1 <- payment.summary %>% 
  pivot_longer(ends_with("pay"), names_to = "Method", values_to = "Payment")   %>% 
  mutate(label = if_else(year == max(year), as.character(Method), NA_character_)) %>% 
  ggplot(mapping = aes(year, Payment, col=Method)) +
  geom_line(show.legend = FALSE) +
  scale_color_discrete(guide = "none") +
  theme_classic() +
  labs(title= "Payment Methods & Amount Paid per Year",
       x= "Year", y="Amount in USD (billion)", caption = "By am_innocenter for #TidyTuesdy 26.11.2019 \n Data source : studentaid.ed.gov")+
  geom_label_repel(aes(label=label),nudge_x = 2, na.rm = TRUE )+
  theme(panel.background = element_rect(fill="gray8")) + #, colour = "#6D9EC1")) +
  theme(text = element_text( size = 12),
        plot.caption = element_text( size=10, color = "grey40"),
        plot.title =element_text(size = 14, face="bold") ) 


```


```{r}

payment2 <- payment.summary %>% 
  pivot_longer(ends_with("pay"), names_to = "Method", values_to = "Payment")   %>% 
  mutate(label = if_else(year == max(year), as.character(Method), NA_character_)) %>% 
  ggplot(mapping = aes(year, Payment, col=Method)) +
  geom_line(show.legend = FALSE) +
  scale_color_discrete(guide ="none") +
  scale_y_log10() +
  theme_classic() +
  labs(title= "Payment Methods & Amount Paid per Year",
       x= "Year", y="Log 10 of Amount in USD (billion)", caption = "By am_innocenter for #TidyTuesdy 26.11.2019 \n Data source : studentaid.ed.gov")+
  geom_label_repel(aes(label=label),nudge_x = 0.4, na.rm = TRUE )+
  theme(panel.background = element_rect(fill="gray8")) + #, colour = "#6D9EC1")) +
  theme(text = element_text( size = 12),
        plot.caption = element_text( size=10, color = "grey40"),
        plot.title =element_text(size = 14, face="bold") ) 
```



```{r}
payment1 + payment2
```

```{r}

loans %>% 
  group_by(agency_name) %>% 
  summarise(Total.starting = sum(starting, na.rm=TRUE)/10^9, 
            Total.added = sum(added, na.rm = TRUE)/10^9,
            Total.repaid = sum(total, na.rm=TRUE)/10^9) %>% 
  arrange(desc(Total.starting)) %>% 
  filter(Total.starting > 30) %>% 
  ungroup() %>% 
  kableExtra::kable() %>% kableExtra::kable_styling()
```



