---
title: "Data Visualisation: Confirmed New Zealand COVID-19 Cases"
description: |
  Short descriptive analysis on the number of COVID-19 cases in New Zealand (NZ)
date: 2020-03-30
draft = true
categories:
  - Data Mining
  - Data viz
  - TidyTuesday
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

> The objective is to provide insights based on the information from the New Zealand (NZ) Ministry of Health. Here is the [link](https://www.health.govt.nz/our-work/diseases-and-conditions/covid-19-novel-coronavirus/covid-19-current-situation/covid-19-current-cases/covid-19-current-cases-details) for obtaining the COVID-19 cases. Will look at the exploring the confirmed cases per District Health Board - commonly known as DHB, gender and age group. 

_With time I might automate the data visualisation - the tricky bit is that the table format and HTML texts/labels keep changing every time. This caused errors in my previous automated visualisation making it difficult to automate data extraction/mining._


```{r, message=FALSE, echo=FALSE}
library(tidyverse)
library(rvest)
library(xml2)
library(lubridate)
library(wordcloud)
library(wordcloud2)
library(tm)
theme_set(theme_bw())
```

## Data Mining and Cleaning 

I used `rvest` and `xml2` packages for data mining, `tidyverse` for data exploration and visualization. The codes can be viewed here.

```{r, echo=F}
url.link <- "https://www.health.govt.nz/our-work/diseases-and-conditions/covid-19-novel-coronavirus/covid-19-current-situation/covid-19-current-cases/covid-19-current-cases-details"

#Extract caption to determine when the COVID-19 cases were last updated
url.link %>%  
  xml2::read_html() %>% 
  html_nodes(xpath = '//table/caption') %>% 
  html_text()-> caption

#extract tables
url.link %>% 
  xml2::read_html() %>% 
  html_nodes("table")  -> tables.cases 
 
#interest is in the confirmed cases, that is table 2
tables.cases[1] %>% 
  html_table() %>% 
  flatten_df -> confirmed.cases

#will look at probable cases later
tables.cases[2] %>% 
  html_table() %>% 
  flatten_df -> probable.cases

#convert to date
confirmed.cases$`Report Date` <- dmy(confirmed.cases$`Report Date`)

#get total cases - this will eb used to calculate proportions
tot <- nrow(confirmed.cases)
```



We have `r tot` confirmed cases `r tail(str_split(caption[1], ",")[[1]],1)`. Below is a glimpse of the confirmed cases - the table only has variables of interest

```{r, echo=F}
glimpse(confirmed.cases[1:6])
```



## Data Visualization 


```{r, echo=F, include=F}
overall.summary <- confirmed.cases %>% 
  count(DHB) %>% #count cases per DHB/District health boards
  arrange(n) %>% 
  mutate(DHB=factor(DHB,DHB), 
         prop=n*100/tot)

overall.summary  %>% 
  filter(n>=5) %>% 
  ggplot(aes(DHB, n, color=DHB, label=paste0(round(prop,0), "%")))  +
    geom_segment(aes(x=DHB, xend=DHB, y=0,yend=n),show.legend = F, size=2) +
    geom_point(size = 12, show.legend = F) +
    coord_flip()+
    labs(x="District Health Board", y="Confirmed Cases", title = caption[1] , subtitle="Where No. of confirmed cases per DHB > 5",
         caption = "Plot by: @am_innocenter \n #Dataviz for #NZ_COVID-19  \n Data source : bit.ly/healthgovtnzCovid")+
    geom_text(show.legend = F, color="black")+
    theme(text = element_text(family = "Comic Sans MS", size = 15),
        plot.caption = element_text( size=10, color = "grey40"),
        plot.title =element_text(size = 16, face="bold", vjust=1) )


```

From the above chart, we observe that Auckland, Southern, Waikato and Waitemata are the top four DHBs which constitutes `r round(sum(overall.summary[17:19, 3]),2)` % of the total confirmed COVID-19 cases in NZ.


When analysing individual variables, I noticed some missing information on gender, age group were missing. We shall exclude these

```{r}
#summary 1 by age and gender
confirmed.cases.by <- confirmed.cases %>% 
  group_by(`Age Group`, Sex) %>% 
  count(`Age Group`) %>% #count cases per DHB
  arrange(`Age Group`) %>% 
  mutate( prop=n*100/tot) %>% 
  filter(! `Age Group`== "" & !Sex =="")

```

Removing them reduced the data set by about `r 
round(1-sum(confirmed.cases.by$n)/nrow(confirmed.cases),2)*100` percent

```{r}
#visualization 1
confirmed.cases.by %>% 
  mutate(count= if_else(Sex=="Female", -n, n)) %>% 
ggplot(mapping= aes(`Age Group`, count, fill=Sex, label=paste0(round(prop,1), "%")))  +
  geom_bar(stat = "identity")+
  coord_flip()+
  scale_y_continuous(breaks = seq(-60, 60, 20),
                     labels = c(seq(60,0,-20), seq(20,60,20)))+
  labs(x="Age group", y="Confirmed Cases", title ="Confirmed cases per Age group and gender" ,subtitle = tail(str_split(caption[1], ",")[[1]],1), caption = "Plot by: @am_innocenter \n #Dataviz for #NZ_COVID-19  \n Data source : bit.ly/healthgovtnzCovid")+
  theme(text = element_text(family = "Comic Sans MS", size = 14),
        plot.caption = element_text( size=10, color = "gray40"),
        plot.title =element_text(size = 16, face="bold", vjust=1) ) +
  geom_text(show.legend = F, color="black", size=4.5)+
  geom_text(aes(3,-45), label="\u2640", hjust = 0.5, size = 25,color = "#F8766D", family = "Comic Sans MS") +
  geom_text(aes(3,30), label="\u2642", hjust = 0.5, size = 25, color = "#00BFC4", family = "Comic Sans MS") 

```

```{r, echo=FALSE}
confirmed.cases %>% 
   group_by(`Age Group`) %>%  count(sort=T) -> cases.age
```

The age group with the highest confirmed cases is `r cases.age[1,1]` which is `r round(cases.age$n[1]*100/tot,2) `% of the total confirmed cases,  followed by `r cases.age[2,1]` with `r round(cases.age$n[2]*100/tot,2) ` % and then `r cases.age[3,1]` with `r round(cases.age$n[3]*100/tot,2) `%. We have more female confirmed cases.




```{r}
#visualization
confirmed.cases %>% 
  filter(! `Age Group`== "" & !Sex =="") %>% 
  group_by(`Report Date`, `Age Group` , Sex) %>% 
  count(`Age Group`) %>% 
  ggplot(mapping=aes(`Report Date`, `Age Group`, size=n , color=Sex)) +
    geom_point() +
    scale_x_date(date_breaks =  "2 days" , date_labels = "%d %b")+
    theme_light()+
    labs(x="", y="Age Group",title = "Confirmed cases trend per Age group and gender" ,subtitle = tail(str_split(caption[1], ",")[[1]],1),
         caption = "Plot by: @am_innocenter \n #Dataviz for #NZ_COVID-19  \n Data source : bit.ly/healthgovtnzCovid")+
    theme(text = element_text(family = "Comic Sans MS", size = 14),
          plot.caption = element_text( size=10, color = "grey40"),
          plot.title =element_text(size = 16, face="bold", vjust=1) )


```

Almost a similar chart as above, this however includes time. The size of the doot indicates the number od confirmed cases recorded. Initially the confirmed cases were from people > 30 years and over time the cases were recorded for people < 20 years including one <1 year and four who are in the 1-4 age group. 

```{r, echo=F, include=F}
##summary 2 by time
confirmed.cases  %>% 
  filter(!Sex =="") %>% 
  group_by(`Report Date`) %>% 
  count(`Report Date`) %>% #count cases per DHB/District health boards
  arrange(n) %>% 
  mutate( prop=n*100/tot) -> total.cases 

total.cases %>% 
  ggplot(mapping=aes(`Report Date`, n, size=n, color="blue", label=n )) +
  geom_point()+
  scale_y_continuous(breaks=seq(0,max(total.cases$n),2))+
  scale_x_date(date_breaks =  "2 days" , date_labels = "%d %b")+
  labs(x="", y="No. of confirmed cases", title = "Confirmed cases trend", subtitle = tail(str_split(caption[1], ",")[[1]],1),
       caption = "Plot by: @am_innocenter \n #Dataviz for #NZ_COVID-19 \n Data source : bit.ly/healthgovtnzCovid")+
  theme(text = element_text(family = "Comic Sans MS", size = 14),
        plot.caption = element_text( size=10, color = "grey40"),
        plot.title =element_text(size = 16, face="bold", vjust=1) )

```

Creating a wordcloud of the Last country before New Zealand. Some of the confirmed cases have New Zealand as the last country to imply within-country travel and this is based on the flight Nos. Hover to get the count for each country.

```{r}
#renaming UK, USA and UAE - they take so much space in the wordcloud
confirmed.cases <- confirmed.cases %>% 
  mutate(LastCountry = 
           case_when(`Last country before NZ` == "United States of America" ~ "USA", 
                    `Last country before NZ` == "United Kingdom" ~ "UK",
                    `Last country before NZ` == "United Arab Emirates" ~ "UAE",
                     TRUE ~ `Last country before NZ`))

#remove spaces for other countries
confirmed.cases$LastCountry <- gsub(" ", "", confirmed.cases$LastCountry)

#creating word corpus
country.corpus <- Corpus(VectorSource(confirmed.cases$LastCountry))
country.corpus <-  TermDocumentMatrix(country.corpus) 

#creating dataframe
m <- as.matrix(country.corpus)
v <- sort(rowSums(m),decreasing=TRUE)
LastCountry.df <- data.frame(word = names(v),freq=v)  

#wordcloud
wordcloud2(LastCountry.df,size=1.6,color='random-dark',  fontFamily="Comic Sans MS")

```

The top five last country of before NZ from the word cloud above and frequent table below:

1. USA       |                    72
2. Australia|               42
3. UAE    |                  31
4. Singapore |                15
5. Qatar |14



