---
title: "Data mining and wrangling : A case of Eid Occurance"
description: |
  Data mining with rvest; case study of Eid observations
categories: 
  - Data Mining
  - Data Viz
date: 2019-06-05
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


Finally, Eid is here! Id-Ul-Fitr commonly known as Eid marks the end of fasting (_Ramadhan_) and is the first day of the Islamic month _Shawwal_. I spent my weekend learning data mining with rvest and analysing data for Eid observations. 



The following [link](https://www.timeanddate.com/holidays/kenya/eid-al-fitr) contains information about Eid, its occurence of per Day, Month, year. For today, I intend to only mine the table and explore. 

### Objectives

1. Mine data - table - from a URL using the package rvest (this was my first time and its really simple)

2. Explore basic data cleaning using dplyr (this is included in tidyverse library) 

3. Explore the occurence of Eid i.e per month, per day

4. Perform visualization using ggplot :-).

This is a learning curve and feel free to drop in your comments and/or suggestions. I will show you a bit of my thought process when analysing and wrangling data.

*Let's go* `r emoji::emoji("smile")` `r emoji::emoji("dancing")` 

_Side note_ check this [link](https://github.com/hadley/emoji) and [this](https://github.com/EmilHvitfeldt/emoji) to learn more about including emojis in a markdown.




Loading packages required

```{r, message =FALSE}
library(tidyverse) # data wrangling
library(rvest) #used for web scraping
library(ggplot2)
library(emoji)
```

## Data Mining 

Loading the data - but first we have to mine it from the URL provided

```{r}

url.page <- read_html('https://www.timeanddate.com/holidays/kenya/eid-al-fitr')

Eid.table <- html_nodes(url.page, 'table')

head(Eid.table) 


```
There are 3 tables - no idea which one contains the Eid data. I will extract all the tables - out of curiosity and they're only 3. If we had several tables - we could explicity use the table names e.g _html_nodes('#table2')_



```{r}
Eid.tables <- url.page %>%
  html_nodes('table') %>% #to select <table> nodes
  .[1:3] %>%
  html_table(fill = TRUE)

str(Eid.tables) # from 2015 t0 2025



```

The second table contains information we are interested in - it has Eid occurance data from the year 2015 - 2025 (some are predictions).


```{r}
Eid <- Eid.tables[[2]]

```
Another method to extact the table is by creating an empty list and populating it with data


```{r}

Eid2 <- list() #creating an empty list

Eid2 <- url.page %>%
  html_nodes('table') %>% 
  html_table(fill = TRUE) %>%
  .[[2]] #populating it with table 2 

```

## EDA and cleaning



```{r}
str(Eid)
```
The data contains (`r dim(Eid)`) - that is 12 observations and 5 variables


In the year 2015 - there exists two entries the second one is the observed and thus will delete the first entry

```{r}
Eid = Eid[-1,]
```


From the structure above, we can see that the names of weekday is written in some language for example Sunday is J2 - I checked the English equivalence in the website and replaced them.

The column names and data types are

```{r}
colnames(Eid)
c(typeof(Eid$Year), typeof(Eid$Weekday), typeof(Eid$Date))
```


The analysis will be based on the month and hence I separate the day from the month in column Date

```{r}
Eid <- Eid %>%
  separate(Date, c('Day', 'Month'))

```


The month of May is written as Mei - I replaced that and the weekday as shown below. 

> From the website: Sun-J2, Mon-J3, Tue-J4, Wed-J5, Thu-Alh, Fri-Ij, Sat-J1

```{r}
Eid$Month <- with(Eid, replace(Month, Month == "Mei", "May"))
Eid$Weekday <- with(Eid, replace(Weekday, Weekday == "J1",  "Sat" ))
Eid$Weekday <- with (Eid, gsub("J2", "Sun", Weekday )) 

```


I had to do these one item after another - I will figure out a way next time maybe a loop



The _replace_ and _gsub_ functions worked for just one item - I tried concatinating the other and got an error while compiling or the matching was not exactly correct. I got this error with gsub
_argument 'replacement' has length > 1 and only the first element will be used_


I also replaced the rest 

```{r, include=FALSE}
Eid$Weekday <- with (Eid, gsub("J3", "Mon", Weekday )) 
Eid$Weekday <- with (Eid, gsub("J4", "Tue", Weekday )) 
Eid$Weekday <- with (Eid, gsub("J5", "Wed", Weekday )) 
Eid$Weekday <- with (Eid, gsub("Alh", "Thu", Weekday )) 
Eid$Weekday <- with (Eid, gsub("Ij", "Fri", Weekday )) 


```

Here is our cleaned data 

```{r}
Eid
```

```{r}

Eid.month <- Eid %>% count(Month, name="Month_occurence")
Eid.month
```

For the past 12 years each month has been represented three times except for July. 
```{r}
Eid.day <- Eid %>% count(Weekday, name='Weekday_occurence')
Eid.day
```
Eid was observed mostly on Monday from the year 2015 - 2025 - this is the future `r emoji::emoji("wink")`

## Visualization

Our fourth objective was to visualize the data and get an insight on the month or day that Eid is observed for the 12 years.

```{r}

ggplot(Eid, aes(Year, Weekday)) +
  geom_point(color="steelblue", shape=20, size =3)+
  scale_x_discrete(limits = c(2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024, 2024, 2025))+ #To preorder the x axis
  labs (title = " Weekday Eid observed per Year")+
  theme_minimal()

```

```{r, layout="l-body-outset", preview=TRUE}
ggplot(Eid, aes(Year, Month))+
  geom_point(aes(colour=Weekday), size = 3)+
  scale_x_discrete(limits = c(2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024, 2024, 2025))+ #To preorder the x axis
  labs(title = "Month Eid observed per year & weekday")+
  theme_minimal()
```






> Take home notes : this was a refresher for me and aluta continua 

I hope you have enjoyed this - though short `r emoji("smile")`


> From this data, the prediction is that next year - Eid will occur on a Monday in the month of May - I shall sit tight and wait for it. 

> Have a blessed Eid 
